---
title: "Bayesian Modelling of Urban Observatory Data"
author: "Jonny Law"
date: "14 August 2019"
output:
  ioslides_presentation:
    logo: figures/NewcastleLogo.svg
    css: my_presentation.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, fig.pos = 'center')
library(tidyverse)
library(latex2exp)
library(ggthemes)
library(leaflet)
library(dlm)
library(jonnylaw)
theme_set(theme_tufte())
```

## The Urban Observatory

> - Sensors stream live data from around Newcastle upon Tyne and Gateshead
> - A variety of different sensor readings such as NO, NO<sub>2</sub>, Temperature and Humidity 
> - Sample sporadically and inaccurately
> - Are not positioned optimally
> - Can we use statistical modelling to determine the true nature of the environmental process 
> - Performing forecasting in time, spatial interpolation and correction of inaccurate readings

## Environment Sensors

```{r, fig.align='center'}
locations = read_csv("data/locations.csv")
## Output a leaflet map of sensor locations
## Remove misspecified sensor locations (with lat & lon = 0.0)
leaflet(locations %>% 
  filter(latitude != 0.0)) %>%
  addTiles() %>%
  addMarkers(~longitude, ~latitude, popup = ~ paste(name, reading, sep = "\n"))
```

## Environment Sensors

```{r read_data, cache=TRUE}
emote = read_csv("data/sensor_data.csv")
```

```{r plot_environment_sensor}
emote %>%
  mutate(Variable = paste0(Variable, " (", Units, ")")) %>%
  ggplot(aes(x = Timestamp, y = Value, colour = Variable)) +
  geom_point(size = 0.1) +
  facet_wrap(~Variable, scales = "free_y") +
  theme(legend.position = "none") +
  labs(title = "Environmental Sensor Readings", 
       subtitle = "Raw readings of sensor data at emote_1108")
```

## Interpolating the Missing Data

```{r gibbs-sampling}
sample_w <- function(theta, mod, shape_w, rate_w) {
  n <- nrow(theta) - 1
  d <- ncol(theta)
  theta_center <- theta[-1, , drop = FALSE] - tcrossprod(theta[-(n + 1), , drop = FALSE], mod$GG)
  SStheta <- drop(sapply(1:d, function(i) crossprod(theta_center[, i])))
  SStheta <- colSums((theta[-1, 1:d, drop = FALSE] - tcrossprod(theta[-(n + 1), , drop = FALSE], mod$GG)[, 1:d])^2)

  1 / rgamma(d, shape = shape_w + 0.5 * (n),
             rate = rate_w + 0.5 * SStheta)
}

center_y <- function(y, theta, mod) 
  y - tcrossprod(theta[-1, , drop = FALSE], mod$FF)

sample_v <- function(y, theta, mod, shape_v, rate_v) {
  p <- ncol(y)
  y_center <- center_y(y, theta, mod)
  SSy <- drop(crossprod(na.omit(y_center))) 
  1 / rgamma(p, shape = shape_v + 0.5 * nrow(y), 
             rate = rate_v + 0.5 * SSy)
}

sample_v_wishart <- function(y, theta, mod, nu, psi) {
  p <- ncol(y)
  y_center <- center_y(y, theta, mod)
  SSy <- drop(crossprod(na.omit(y_center))) 

  dof <- nu + n
  scale <- psi + SSy

  MCMCpack::riwish(1, v = dof, S = scale)
}

dlm_gibbs <- function(y, mod, theta, shape_v, rate_v, shape_w, rate_w, m) {
  p <- nrow(mod(theta)$FF)
  d <- nrow(mod(theta)$GG)
  v <- matrix(NA_real_, nrow = m, ncol = p)
  w <- matrix(NA_real_, nrow = m, ncol = d)
  v[1,] <- theta[1:p]
  w[1,] <- theta[(p+1):(p+d)]
  theta <- c(w[1,], v[1,])
  for (i in seq_len(m)) {
    filtered <- dlmFilter(y, mod(theta))
    sampled <- dlmBSample(filtered)
    w[i,] <- sample_w(sampled, mod(theta), shape_w, rate_w)
    v[i,] <- sample_v(y, sampled, mod(theta), shape_v, rate_v)
    theta <- c(w[i,], v[i,])
  }
  out <- cbind(w, v) %>% as_tibble()
  colnames(out) <- c(paste0("w", seq_len(d)), paste0("v", seq_len(p)))
  out %>% 
    add_column(iteration = seq_len(m))
}
```

 - Centre and scale the data (subtract sample mean, divide by sample standard deviation)

```{r}
rounded_data <- emote %>% 
  filter(Variable %in% c("Temperature", "Humidity")) %>% 
  spread(key = Variable, value = Value) %>% 
  select(timestamp = Timestamp, Humidity, Temperature) %>% 
  mutate(timestamp = lubridate::round_date(timestamp, unit = "1 hour")) %>% 
  group_by(timestamp) %>% 
  summarise_at(vars(Humidity, Temperature), mean, na.rm = T)

emote <- rounded_data %>% 
  expand(timestamp = seq(from = lubridate::round_date(min(timestamp), "1 hour"), to = lubridate::round_date(max(timestamp), "1 hour"), by = "1 hour")) %>% 
  left_join(rounded_data) %>% 
  mutate(mean_humidity = mean(Humidity, na.rm = T), sd_humidity = sd(Humidity, na.rm = T),
         mean_temperature = mean(Temperature, na.rm = T), sd_temperature = sd(Temperature, na.rm = T)) %>% 
  mutate(norm_humidity = (Humidity - mean_humidity) / sd_humidity, 
         norm_temperature = (Temperature - mean_temperature) / sd_temperature)

emote %>% 
  gather(key, value, norm_humidity, norm_temperature) %>% 
  ggplot(aes(x = timestamp, y = value, colour = key)) +
  geom_line() +
  facet_wrap(~key, ncol = 2)
``` 

## Fit a Parametric Model

$$\begin{align}
y_t &= Fx_t + v_t, &v_t &\sim \mathcal{N}(0, V), \\
x_t &= Gx_{t-1} + w_t, &w_t &\sim \mathcal{N}(0, W), \\
x_0 &\sim \mathcal{N}(m_0, C_0).
\end{align}$$

```{r model}
multivariate_model <- function(theta) {
  (dlmModPoly(order = 1, dV = theta[1], dW = theta[3]) + 
    dlmModTrig(s = 24, q = 3, dV = 0, dW = theta[4:9])) %+%
  (dlmModPoly(order = 1, dV = theta[2], dW = theta[10]) + 
    dlmModTrig(s = 24, q = 3, dV = 0, dW = theta[11:16]))
}
```

```{r create-matrix}
ys <- emote %>% 
  select(norm_humidity, norm_temperature) %>% 
  as.matrix()
```

```{r mle, eval=FALSE}
fit <- dlmMLE(ys, parm = rep(1, times = 16), build = multivariate_model)
write_rds(fit, "models/mle_fit.Rds")
```

```{r}
fit <- read_rds(path = "models/mle_fit.Rds")
```

```{r chains, eval=FALSE}
future::plan(future::multiprocess)
temp_humid_chains <- furrr::future_map_dfr(
  .x = 1:2,
  .f = function(x)
    dlm_gibbs(
      y = ys,
      mod = multivariate_model,
      theta = fit$par,
      shape_v = 2.0,
      rate_v = 3.0,
      shape_w = 2.0,
      rate_w = 3.0,
      m = 5e3
    ),
  .id = "chain"
)
write_rds(x = temp_humid_chains, "models/temp_humid_chains.Rds")
```

```{r}
temp_humid_chains <- read_rds(path = "models/temp_humid_chains.Rds")
```

```{r}
chains <- 
  temp_humid_chains %>% 
    filter(iteration > 1000) %>% 
    thin(10) %>% 
    gather(key = Parameter, value, -chain, -iteration) 

chains %>% 
    ggplot2::ggplot() +
    ggplot2::geom_density(ggplot2::aes(x = value, fill = as.factor(chain)), alpha = 0.5) +
    ggforce::facet_wrap_paginate(~Parameter, scales = "free", nrow = 3, ncol = 3, strip.position = "right", page = 1) +
    ggplot2::theme(legend.position = "none")
```

-------

```{r}
chains %>% 
    ggplot2::ggplot() +
    ggplot2::geom_density(ggplot2::aes(x = value, fill = as.factor(chain)), alpha = 0.5) +
    ggforce::facet_wrap_paginate(~Parameter, scales = "free", nrow = 3, ncol = 3, strip.position = "right", page = 2) +
    ggplot2::theme(legend.position = "none")
```

## One Step Forecasting using the Kalman Filter

```{r interpolation-filter}
theta_mean <- temp_humid_chains %>% 
  summarise_at(vars(2:17), mean) %>% 
  .[1,]
filtered <- dlmFilter(y = ys, mod = multivariate_model(theta_mean))
f <- multivariate_model(theta_mean)$FF
v <- multivariate_model(theta_mean)$V
rt <- dlmSvd2var(u = filtered$U.R, d = filtered$D.R)
qt <- map(rt, ~ f %*% . %*% t(f) + v)

humid <- emote %>% 
  add_column(pred_humid = filtered$f[,1],
             variance_humid = qt %>% map(~ .[1,1]) %>% unlist()) %>% 
  filter(timestamp > "2017-09-01") %>% 
  mutate(upper = qnorm(p = 0.975, mean = pred_humid, sd = sqrt(variance_humid)),
         lower = qnorm(p = 0.025, mean = pred_humid, sd = sqrt(variance_humid))) %>% 
  gather(key, value, pred_humid, norm_humidity) %>% 
  ggplot(aes(x = timestamp, y = value, colour = key)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), colour = NA, alpha = 0.3) +
  geom_line() +
  coord_cartesian(ylim = c(-10, 10))

temp <- emote %>% 
  add_column(pred_temp = filtered$f[,2],
             variance_temp = qt %>% map(~ .[2,2]) %>% unlist()) %>% 
    filter(timestamp > "2017-09-01") %>% 
  mutate(upper = qnorm(p = 0.975, mean = pred_temp, sd = sqrt(variance_temp)),
       lower = qnorm(p = 0.025, mean = pred_temp, sd = sqrt(variance_temp))) %>% 
  gather(key, value, pred_temp, norm_temperature) %>% 
  ggplot(aes(x = timestamp, y = value, colour = key)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), colour = NA, alpha = 0.3) +
  geom_line() +
  coord_cartesian(ylim = c(-19, 10))

gridExtra::grid.arrange(humid, temp)
```

## Interpolation using Smoothing

```{r smoothed-interpolation}
smoothed <- dlmSmooth(filtered)
smoothed_states <- smoothed$s
pred <- apply(smoothed_states, 1, function(x) f %*% x)
rt <- dlmSvd2var(u = smoothed$U.S, d = smoothed$D.S)
pred_var <- map(rt, ~ f %*% . %*% t(f) + v)

humid <- emote %>% 
  add_column(pred_humid = pred[1,-1],
             variance_humid = pred_var %>% map(~ .[1,1]) %>% unlist() %>% .[-1]) %>% 
  filter(timestamp > "2017-09-01") %>% 
  mutate(upper = qnorm(p = 0.975, mean = pred_humid, sd = sqrt(variance_humid)),
         lower = qnorm(p = 0.025, mean = pred_humid, sd = sqrt(variance_humid))) %>% 
  gather(key, value, pred_humid, norm_humidity) %>% 
  ggplot(aes(x = timestamp, y = value, colour = key)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), colour = NA, alpha = 0.3) +
  geom_line() +
  coord_cartesian(ylim = c(-10, 10))

temp <- emote %>% 
  add_column(pred_temp = pred[2,-1],
           variance_temp = pred_var %>% map(~ .[1,1]) %>% unlist() %>% .[-1]) %>% 
  mutate(upper = qnorm(p = 0.975, mean = pred_temp, sd = sqrt(variance_temp)),
       lower = qnorm(p = 0.025, mean = pred_temp, sd = sqrt(variance_temp))) %>%   
  filter(timestamp > "2017-09-01") %>% 
  gather(key, value, pred_temp, norm_temperature) %>% 
  ggplot(aes(x = timestamp, y = value, colour = key)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), colour = NA, alpha = 0.3) +
  geom_line() +
  coord_cartesian(ylim = c(-10, 10))

gridExtra::grid.arrange(humid, temp)
```